Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
a few more edits in the intro

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed or polymorphic lambda calculus, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a full universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs
  for dependent type theories} Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code. As a result, their use of logical relations are
difficult to isolate from their surrounding contexts, and thus, not accessible
to casual readers.

@@

The result of our work is an artifact that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set-theoretic notation and terminology so
that the material is accessible to readers with a general mathematical
background.  That said, our explanations do not stray too far away from our
proof scripts.  We link each lemma directly to its counterpart in the proof
script, anticipating that readers may wish to reference and validate our
formalism.  The typeset proofs purposefully follow the mechanized proofs
closely while avoiding, as much as possible, artifacts specific to Coq.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relations is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed or polymorphic lambda calculus, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a full universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs
  for dependent type theories?} Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code. As a result, their use of logical relations are
difficult to isolate from their surrounding contexts and inaccessible
to casual readers.

@@

The result of our work is an artifact that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set-theoretic notation and terminology so
that the material is accessible to readers with a general mathematical
background.  That said, our explanations do not stray too far away from our
proof scripts.  We link each lemma directly to its counterpart in the proof
script, anticipating that readers may wish to see how these results may be
expressed and verified in a proof assistant.  The typeset proofs purposefully
follow the mechanized proofs closely while avoiding, as much as possible,
artifacts specific to Coq.
